---
title: "Pokemon GO data analysis"
author: "Levi Waldron"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the Pokemon dataset

Pokemon dataset from https://www.openintro.org/stat/data/?data=pokemon.

Option 1, after downloading the csv file to the same directory as this script (PokemonGO.Rmd):

```{r, eval=FALSE}
pokemon <- read.csv("pokemon.csv")
```

Option 2, downloading from the internet:

```{r}
pokemon = read.csv("https://www.openintro.org/stat/data/pokemon.csv")
```

# Linear models on Combat Power

```{r}
pfit = lm(cp ~ species + hp + weight + height, data=pokemon)
pfit2 = lm(cp ~ (species + hp + weight + height)^2, data=pokemon) #all 2-way interactions
pfit3 = lm(cp ~ . - notes - name - attack_weak, data=pokemon)  # all available predictors except for notes, names, and attack_weak.
```

# `dplyr` introduction

Help at https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

Create a dataset that excludes all columns whose name ends in "new" (notice the - before `ends_with()`:

```{r}
library(dplyr)
pokemon2 = select(pokemon, -ends_with("new"))
```

Exclude columns whose name ends in new, then keep only rows where the species is "Pidgey":

```{r}
pokemon2 = pokemon %>% 
  select(-ends_with("new")) %>%
  filter(species == "Pidgey")
summary(pokemon$species)
summary(pokemon2$species)
```

# `ggplot2` introduction

Help at https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

## A blank canvas
`ggplot2` plots are built up piece by piece. The following creates a blank plot, and store all the data to make that plot and build on it in the `myplot` object:

```{r}
library(ggplot2)
myplot = ggplot(aes(cp, cp_new), data=pokemon)
```

## A scatterplot
To add a scatterplot with colors and a legend, just add a `geom_point()` call. We could save this to a new object, like `myplot2`, but in this example we won't create any new object, just make a scatterplot:

```{r}
myplot + geom_point(aes(color=species))
```

## A scatterplot plus regression lines
Try doing the following one by one, just adding new things to the existing plot. These functions are all documented in the `ggplot2` cheat sheet. You can also try skipping some of the lines. Note that nothing will happen until there is a line that doesn't end in "+"

```{r}
ggplot(aes(cp, cp_new), data=pokemon) +
  geom_point(aes(color=species)) +  #scatterplot
  geom_smooth(method="lm") +  #linear regression line and confidence bands
  theme_bw()  #get rid of the grey background
```

## Boxplot

Now let's make a boxplot:
```{r}
ggplot(aes(species, cp), data=pokemon) +
  geom_boxplot(fill="grey") +
  ggtitle("Combat Power by Species") +
  xlab("Species") + 
  ylab("Combat Power") +
  theme_bw()
```

Try other kinds of plots, `geom_violin` and `geom_dotplot`...
