
---
title: "BIOS621 / 821 Session 5"
author: "Levi Waldron"
date: "loglinear regression part 2"
output: slidy_presentation
---

Welcome and outline - session 5
=================================================
* Review of log-linear Poisson glm
* Review of diagnostics and interpretation of coefficients
* Over-dispersed models:
    + negative binomial distribution
* Zero-inflated models
* Vittinghoff section 8.1-8.3

Components of GLM
========================================================

* **Random component** specifies the conditional distribution for the response variable - it doesn’t have to be normal but can be any distribution that belongs to the “exponential” family of distributions
* **Systematic component** specifies linear function of predictors (linear predictor)
* **Link** [denoted by g(.)] specifies the relationship between the expected value of the random component and the systematic component, can be linear or nonlinear  


Motivating example: Choice of Distribution
=================================================
* Count data are often modeled as Poisson distributed:
       + mean $\lambda$ is greater than 0
       + variance is also $\lambda$
       + Probability density $P(k, \lambda) = \frac{\lambda^k}{k!} e^{-\lambda}$
       
```{r, echo=FALSE}
##par(cex=2)  #increase size of type and axes
plot(x=0:10, y=dpois(0:10, lambda=1), 
     type="b", lwd=2,
     xlab="Counts (k)", ylab="Probability density")
lines(x=0:10, y=dpois(0:10, lambda=2), 
      type="b", lwd=2, lty=2, pch=2)
lines(x=0:10, dpois(0:10, lambda=4), 
      type="b", lwd=2, lty=3, pch=3)
legend("topright", lwd=2, lty=1:3, pch=1:3,
       legend=c(expression(paste(lambda, "=1")),
                expression(paste(lambda, "=2")),
                expression(paste(lambda, "=4"))))
```

Poisson model: the GLM
=================================================

The **systematic part** of the GLM is:
$$
log(\lambda_i) = \beta_0 + \beta_1 \textrm{RACE}_i + \beta_2 \textrm{TRT}_i + \beta_3 \textrm{ALCH}_i + \beta_4 \textrm{DRUG}_i
$$
Or alternatively:
$$
\lambda_i = exp \left( \beta_0 + \beta_1 \textrm{RACE}_i + \beta_2 \textrm{TRT}_i + \beta_3 \textrm{ALCH}_i + \beta_4 \textrm{DRUG}_i \right)
$$

The **random part** is (Recall the $\lambda_i$ is both the mean and variance of a Poisson distribution):
$$
y_i \sim Poisson(\lambda_i)
$$


Example: Risky Drug Use Behavior
=================================================
* Download the "needle_sharing" dataset in csv format
* Outcome is # times the drug user shared a syringe in the past month (shared_syr)
* Predictors: sex, ethn, homeless

\small
```{r}
needledat = read.csv("needle_sharing.csv")
needledat2 <- needledat[needledat$sex %in% c("M", "F") & 
    needledat$ethn %in% c("White", "AA", "Hispanic"), ]
summary(needledat2$shared_syr)
var(needledat2$shared_syr, na.rm=TRUE)
```

Example: Risky Drug Use Behavior
=================================================
```{r, echo=FALSE}
par(mfrow=c(1,2))
##par(cex=2)
hist(needledat2$shared_syr, main="")
plot(sort(needledat2$shared_syr), pch=".")
```

* There are a _lot_ of zeros and variance is much greater than mean
    + Poisson model is probably not a good fit

Risky Drug Use Behavior: fitting a Poisson model
=================================================

```{r, echo=FALSE}
needledat2$homeless <- factor(needledat2$homeless, levels=0:1, 
                             labels=c("no", "yes"))
```

```{r, results='hide', cache=TRUE}
fit.pois <- glm(shared_syr ~ sex + ethn + homeless, 
           data=needledat2, family=poisson(link="log"))
```

Risky Drug Use Behavior: residuals plots
=================================================
```{r, echo=FALSE, warning=FALSE}
par(mfrow=c(2,2))
plot(fit.pois)
```
* Poisson model is definitely not a good fit.

When the Poisson model doesn't fit
=================================================
* inference from log-linear models is sensitive to assumptions on the distribution of residuals (e.g. Poisson)
* In the Poisson distribution, the variance is equal to the mean.
* _i.e._ if subjects with a particular pattern of covariates have a mean of 4 visits/yr, then variance is also 4 and the standard deviation is 2 visits / yr.
* The Poisson distribution often fails when the variance exceeds the mean
    + You can _check_ this assumption      
* Can use alternative random distributions:
    + Negative binomial distribution
* Can introduce zero-inflation

Negative binomial distribution
=================================================
* The binomial distribution is the number of successes in n trials:
    + Roll a die ten times, how many times do you see a 6?
* The negative binomial distribution is the number of successes it takes to observe r failures:
    + How many times do you have to roll the die to see a 6 ten times?
    + Note that the number of rolls is no longer fixed.
    + In this example, p=5/6 and a 6 is a "failure"

Negative binomial GLM
=================================================

*One way* to parametrize a NB model is with a **systematic part** equivalent to the Poisson model:
$$
log(\lambda_i) = \beta_0 + \beta_1 \textrm{RACE}_i + \beta_2 \textrm{TRT}_i + \beta_3 \textrm{ALCH}_i + \beta_4 \textrm{DRUG}_i
$$
Or:
$$
\lambda_i = exp \left( \beta_0 + \beta_1 \textrm{RACE}_i + \beta_2 \textrm{TRT}_i + \beta_3 \textrm{ALCH}_i + \beta_4 \textrm{DRUG}_i \right)
$$

And a **random part**:
$$
y_i \sim NB(\lambda_i, \theta)
$$

* $\theta$ is a **dispersion parameter** that is estimated
* When $\theta = 0$ it is equivalent to Poisson model
* `MASS::glm.nb()` uses this parametrization, `dnbinom()` does not
* The Poisson model can be considered **nested** within the Negative Binomial model

Negative Binomial Random Distribution
=================================================
```{r, echo=FALSE}
plot(x=0:40, y=dnbinom(0:40, size=10, prob=0.5), 
     type="b", lwd=2, ylim=c(0, 0.2),
     xlab="Counts (k)", ylab="Probability density")
lines(x=0:40, y=dnbinom(0:40, size=20, prob=0.5), 
      type="b", lwd=2, lty=2, pch=2)
lines(x=0:40, y=dnbinom(0:40, size=10, prob=0.3),
      type="b", lwd=2, lty=3, pch=3)
legend("topright", lwd=2, lty=1:3, pch=1:3,
       legend=c("n=10, p=0.5", "n=20, p=0.5", "n=10, p=0.3"))
```

Compare Poisson vs. Negative Binomial
=================================================
Negative Binomial Distribution has two parameters: # of trials n, and probability of success p

```{r, echo=FALSE}
plot(x=0:40, y=dnbinom(0:40, size=10, prob=0.5), 
     type="b", lwd=2, ylim=c(0, 0.15),
     xlab="Counts (k)", ylab="Probability density")
lines(x=0:40, y=dnbinom(0:40, size=20, prob=0.5), 
      type="b", lwd=2, lty=2, pch=2)
lines(x=0:40, y=dnbinom(0:40, size=10, prob=0.3),
      type="b", lwd=2, lty=3, pch=3)
lines(x=0:40, y=dpois(0:40, lambda=9), col="red")
lines(x=0:40, y=dpois(0:40, lambda=20), col="red")
legend("topright", lwd=c(2,2,2,1), lty=c(1:3,1), pch=c(1:3,-1), col=c(rep("black", 3), "red"),
       legend=c("n=10, p=0.5", "n=20, p=0.5", "n=10, p=0.3", "Poisson"))
```

Risky drug behavior: Negative Binomial Regression
=================================================
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(MASS)
fit.negbin <- glm.nb(shared_syr ~ sex + ethn + homeless, 
                     data=needledat2)
summary(fit.negbin)
```

Likelihood ratio test 
=================================================

Recall from class 2 the Deviance:

$\Delta (\textrm{D}) = -2 * \Delta (\textrm{log likelihood})$

And recall the difference in deviance under $H_0$ (no improvement in fit) is *chi-square distributed*, with df equal to the difference in df of the two models:

```{r}
(ll.negbin <- logLik(fit.negbin))
(ll.pois <- logLik(fit.pois))
pchisq(2 * (ll.negbin - ll.pois), df=1, lower.tail=FALSE)
```


Risky Drug Use Behavior: NB regression residuals plots
=================================================
```{r, echo=FALSE, warning=FALSE}
par(mfrow=c(2,2))
plot(fit.negbin)
```

Zero Inflation
=================================================
* Two-step model:
    1. logistic model to determine whether count is zero or Poisson/NB
    2. Poisson or NB regression distribution for $y_i$ not set to zero by *1.*

Poisson Distribution with Zero Inflation
=================================================
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(gamlss)
##par(cex=2)  #increase size of type and axes
plot(x=0:10, y=dpois(0:10, lambda=2), 
     type="b", lwd=2, ylim=c(0, 0.5),
     xlab="Counts (k)", ylab="Probability density")
lines(x=0:10, y=dZIP(0:10, mu=2, sigma=0.2),
      type="b", lwd=2, lty=2, pch=2)
lines(x=0:10, y=dZIP(0:10, mu=2, sigma=0.4),
      type="b", lwd=2, lty=3, pch=3)
legend("topright", lwd=2, lty=1:3, pch=1:3,
       legend=c(expression(paste(lambda, "=2")),
                expression(paste("ZIP: ", lambda, "=2, ", Sigma, "=0.2")),
                expression(paste("ZIP: ", lambda, "=2, ", Sigma, "=0.4"))))
```

Risky drug behavior: Zero-inflated Poisson regression
=================================================
```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
library(pscl)
fit.ZIpois <- zeroinfl(shared_syr ~ sex+ethn+homeless, 
                        dist="poisson",data=needledat2)
```

Zero-inflated Poisson regression - the model
=================================================
```{r}
summary(fit.ZIpois)
```


Risky drug behavior: Zero-inflated Negative Binomial regression
=================================================
```{r, echo=TRUE, results='hide', message=FALSE}
fit.ZInegbin <- zeroinfl(shared_syr ~ sex+ethn+homeless, 
                        dist="negbin", data=needledat2)
```

* *NOTE*: zero-inflation model can include any of your variables as predictors
* *WARNING* Default in `zerinfl()` function is to use _all_ variables as predictors in logistic model 

Zero-inflated Negative Binomial regression - model 1
=================================================

\tiny
```{r}
summary(fit.ZInegbin)
```

Zero-inflated Negative Binomial regression - simplified ZI model
=================================================
* Model is much more interpretable if the exposure of interest is _not_ included in the zero-inflation model.
* E.g. with HIV status as the only predictor in zero-inflation model:

```{r, echo=TRUE, results='hide', message=FALSE}
needledat2$hiv <- factor(ifelse(needledat2$hivstat==0,
                                "negative","positive"))
fit.ZInb2<-zeroinfl(shared_syr~sex+ethn+homeless+hiv|hiv, 
                        dist="negbin", data=needledat2)
```

Zero-inflated Negative Binomial regression - model 2
=================================================

\tiny
```{r}
summary(fit.ZInb2)
```


Intercept-only zero-inflation model
=================================================

\tiny
```{r}
fit.ZInb3 <- zeroinfl(shared_syr~sex+ethn+homeless|1, 
                        dist="negbin", data=needledat2)
summary(fit.ZInb3)
```


Residuals vs. fitted values
=================================================
I invisibly define functions plotpanel1 and plotpanel2 that will work for all types of models (see .R or .Rmd file for functions).  These use Pearson residuals.
```{r, echo=FALSE}
plotpanel1 <- function(fit, ...){
  plot(x=predict(fit), y=residuals(fit, type="pearson"), 
       xlab="Predicted Values", ylab="Pearson Residuals", ...)
  abline(h=0, lty=3)
  lines(lowess(x=predict(fit), y=resid(fit, type="pearson")), 
        col="red")
}
plotpanel2 <- function(fit, ...){
  resids <- scale(residuals(fit, type="pearson"))
  qqnorm(resids, ylab="Std Pearson resid.", ...)
  qqline(resids)
}
```

```{r, echo=FALSE, warning=FALSE}
par(mfrow=c(2,2))
plotpanel1(fit.pois, main="Residuals vs. Fitted\n Poisson")
plotpanel1(fit.negbin, main="Residuals vs. Fitted\n Negative Binomial")
plotpanel1(fit.ZIpois, main="Residuals vs. Fitted\n Zero-inflated Poisson")
plotpanel1(fit.ZInegbin, main="Residuals vs. Fitted\n Zero-inflated Negative Binomial")
```


Quantile-quantile plots for residuals
=================================================
```{r, echo=FALSE, warning=FALSE}
par(mfrow=c(2,2))
plotpanel2(fit.pois, main="Normal Q-Q Plot\n Poisson")
plotpanel2(fit.negbin, main="Normal Q-Q Plot\n Negative Binomial")
plotpanel2(fit.ZIpois, main="Normal Q-Q Plot\n Zero-inflated Poisson")
plotpanel2(fit.ZInegbin, main="Normal Q-Q Plot\n Zero-inflated Negative Binomial")
```
_still_ over-dispersed - ideas?

Inference from the different models
=================================================
\tiny
```{r, echo=FALSE, results="asis", message=FALSE}
library(stargazer)
stargazer(fit.pois, fit.negbin, fit.ZIpois, fit.ZInegbin, fit.ZInb3, 
          title="", type="html",
          style="default", header=FALSE)
```
Zero-inflated models are 3) Poisson, 4) Negative Binomial, and 5) Negative Binomial with intercept-only zero inflation model.

Example of plotting observed and predicted counts
=================================================

```{r, echo=FALSE}
observed <- data.frame(table(needledat2$shared_syr))
fit.lm <- lm(shared_syr~sex+ethn+homeless, data=needledat2)
pred.lm <- data.frame(table(round(predict(fit.lm))))
pred.pois <- data.frame(table(round(predict(fit.pois, type="response"))))
pred.negbin <- data.frame(table(round(predict(fit.negbin, type="response"))))
pred.ZIpois <- data.frame(table(round(predict(fit.ZIpois, type="response"))))
pred.ZInegbin <- data.frame(table(round(predict(fit.ZInegbin, type="response"))))

plot(as.numeric(observed[, 1]), observed[, 2], type="l", lwd=2,
     xlab="Counts", ylab="Frequency")
lines(as.numeric(pred.pois[, 1]), pred.pois[, 2], lty=2, lwd=2)
lines(as.numeric(pred.negbin[, 1]), pred.negbin[, 2], lty=3, lwd=2)
lines(as.numeric(pred.ZInegbin[, 1]), pred.ZInegbin[, 2], lty=4, lwd=2)
legend("topright", legend=c("Observed", "Poisson", "Negbin", "ZI negbin"),
       lty=1:4, lwd=2)
```

Lab exercises
=================================================

* Perform chi-square nested deviance tests for zero-inflated models
* Try fitting the needle dataset using a zero-inflated gamma count distribution

Resources for R and SAS
=================================================
* Short, practical tutrorials on regression in R and SAS from UCLA at http://www.ats.ucla.edu/stat/:
    + Poisson Regression: http://www.ats.ucla.edu/stat/r/dae/poissonreg.htm
    + Negative Binomial: http://www.ats.ucla.edu/stat/r/dae/nbreg.htm
    + Zero-inflated Poisson: http://www.ats.ucla.edu/stat/r/dae/zipoisson.htm
    + Zero-inflated Negative Binomial: http://www.ats.ucla.edu/stat/r/dae/zinbreg.htm
    
