---
title: "BIOS621 Session 2"
author: "Levi Waldron"
output: beamer_presentation
---

Welcome and outline - session 2
========================================================

* brief overview of multiple regression (Chapter 4)
* Linear Regression as a Generalized Linear Model (Chapter 5)
* Statistical inference for logistic regression

Learning objectives - session 2
========================================================

* define generalized linear models (GLM)
* define linear and logistic regression as special cases of GLMs
* distinguish between additive and multiplicative models
* define Pearson and deviance residuals
* additional familiarity with R, including `dplyr` and `ggplot2`

Multiple Linear Regression Model
========================================================
Systematic component:

$$
E[y|x] = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \beta_p x_p
$$

- $x_p$ are the predictors or independent variables
- $y$ is the outcome, response, or dependent variable
- $E[y|x]$ is the expected value of $y$ given $x$
- $\beta_p$ are the regression coefficients

Multiple Linear Regression Model
========================================================

Systematic plus random component:

$y_i = E[y|x] + \epsilon_i$

$y_i = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \beta_p x_p + \epsilon_i$

Assumption: $\epsilon_i \stackrel{iid}{\sim} N(0, \sigma_\epsilon^2)$

* Normal distribution
* Mean zero at every value of predictors
* Constant variance at every value of predictors
* Values that are statistically independent

Generalized Linear Models
========================================================
* Linear regression is a special case of a broad family of models called “Generalized Linear Models” (GLM)
* This unifying approach allows to fit a large set of models using maximum likelihood estimation methods (MLE) (Nelder & Wedderburn, 1972)
* Can model many types of data directly using appropriate distributions, e.g. Poisson distribution for count data
* Transformations of $Y$ not needed

Components of GLM
========================================================

* **Random component** specifies the conditional distribution for the response variable
    + doesn’t have to be normal
    + can be any distribution in the "exponential" family of distributions
* **Systematic component** specifies linear function of predictors (linear predictor)
* **Link** [denoted by g(.)] specifies the relationship between the expected value of the random component and the systematic component
    + can be linear or nonlinear  

Linear Regression as GLM
========================================================

* **The model**: $y_i = E[y|x] + \epsilon_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ... + \beta_p x_{pi} + \epsilon_i$

* **Random component** of $y_i$ is normally distributed:   $\epsilon_i \stackrel{iid}{\sim} N(0, \sigma_\epsilon^2)$

* **Systematic component** (linear predictor): $\beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ... + \beta_p x_{pi}$

* **Link function** here is the _identity link_: $g(E(y | x)) = E(y | x)$.  We are modeling the mean directly, no transformation.

Logistic Regression as GLM
========================================================

* **The model**: 
$$
Logit(P(x)) = log \left( \frac{P(x)}{1-P(x)} \right) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ... + \beta_p x_{pi}
$$

* **Random component**: $y_i$ follows a Binomial distribution (outcome is a binary variable)

* **Systematic component**: linear predictor 
$$
\beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ... + \beta_p x_{pi}
$$

* **Link function**: _logit_ (log of the odds that the event occurs)

$$
g(P(x)) = logit(P(x)) = log\left( \frac{P(x)}{1-P(x)} \right)
$$

$$
P(x) = g^{-1}\left( \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ... + \beta_p x_{pi}
 \right)
$$

logit function
=================================================
\small
```{r, fig.height=6}
logit <- function(P) log(P/(1-P))
plot(logit, xlab="Probability", ylab="Log-odds",
     cex.lab=1.5, cex.axis=1.5)
```

Inverse logit function
=================================================
```{r}
invLogit <- function(x) 1/(1+exp(-x))
```

```{r, fig.height=6, echo=FALSE}
plot(invLogit, xlim=c(-6, 6), cex.lab=1.5, cex.axis=1.5,
     xlab="Log-odds", ylab="Probability")
```

Assumptions in logistic regression
=================================================
1. `y_i` follows a Binomial distribution.
2. The mean `E(y|x) = P(x)` is given by the logit function.
3. Values of the outcome are statistically independent.

Additive vs. Multiplicative models
=================================================

* Linear regression is an _additive_ model
    + _e.g._ for two binary variables $\beta_1 = 1.5$, $\beta_2 = 1.5$.
    + If $x_1=1$ and $x_2=1$, this adds 3.0 to $E(y|x)$
* Logistic regression is a _multiplicative_ model
    + If $x_1=1$ and $x_2=1$, this adds 3.0 to $log(\frac{P}{1-P})$
    + Odds-ratio $\frac{P}{1-P}$ increases 20-fold: $exp(1.5+1.5)$ or $exp(1.5) * exp(1.5)$

Motivating example: contraceptive use data
=================================================
From http://data.princeton.edu/wws509/datasets/#cuse

```{r, echo=FALSE}
cuse <- read.table("http://data.princeton.edu/wws509/datasets/cuse.dat", header=TRUE)
summary(cuse)
```

Motivating example: contraceptive use data
=================================================
No interactions:

\tiny
```{r}
fit1 <- glm(cbind(using, notUsing) ~ age + education + wantsMore, 
           data=cuse, family=binomial("logit"))
summary(fit1)
```

Pearson residuals for logistic regression
=================================================
Take the difference between observed and fitted values (on probability scale 0-1), and divide by the standard deviation of the observed value.

* Let $\hat y_i$ be the best-fit predicted probability for each data point, i.e. $g^{-1}(\beta_0 + \beta_1 x_{1i} + ...)$
* $y_i$ is the observed value, either 0 or 1.

$$
r_i = \frac{y_i - \hat y_i}{ \sqrt{ \frac{\hat y_i(n_i - \hat y_i)}{n_i} }}
$$

Summing the squared Pearson residuals produces the _Pearson Chi-squared statistic_:

$$
\chi ^2 = \sum_i r_i^2
$$

Deviance residuals for logistic regression
=================================================

* Deviance residuals and Pearson residuals converge for high degrees of freedom
* Deviance residuals indicate the contribution of each point to the model _likelihood_
* Definition of deviance residuals:

$$
d_i = s_i \sqrt{ -2 ( y_i \log \hat y_i + (1-y_i) \log (1 - \hat y_i) ) }
$$

Where $s_i = 1$ if $y_i = 1$ and $s_i = -1$ if $y_i = 0$.

* Summing the deviances gives the overall deviance: $D = \sum_i d_i^2$

Model likelihood and deviance
=================================================

* The _likelihood_ of a model is the probability of the observed outcomes given the model, sometimes written as:
    + $L(\theta | data) = P(data|\theta)$.
* Deviance residuals and the difference in log-likelihood between two models are related by:

$\Delta (\textrm{D}) = -2 * \Delta (\textrm{log likelihood})$

Likelihood Ratio Test
=================================================

* Use to assess whether the reduction in deviance provided by a more complicated model indicates a better fit
* It is equivalent of the nested Analysis of Variance is a nested Analysis of Deviance
* The difference in deviance under $H_0$ is *chi-square distributed*, with df equal to the difference in df of the two models.


Likelihood Ratio Test (cont'd)
=================================================

\scriptsize
```{r}
fit0 <- glm(cbind(using, notUsing) ~ -1, data=cuse, 
            family=binomial("logit"))
anova(fit0, fit1, test="LRT")
```

Wald test for individual regression coefficients
=================================================

* Can use partial Wald test for a single coefficient: 
    + $\frac{\hat{\beta}}{\sqrt{var(\hat{\beta)}}} \sim N(0, 1)$
    + $\frac{\hat{\beta}^2 }{var(\hat{\beta)}} \sim   \chi^2_{df=1}$ (large sample)
* Wald CI for $\beta$: $\hat{\beta} \pm t_{1-\alpha/2, n-1} \sqrt{var(\hat{\beta})}$
* Wald CI for odds-ratio: $e^{\hat{\beta} \pm t_{1-\alpha/2, n-1} \sqrt{var(\hat{\beta})}}$

_Note_: Wald test confidence intervals on coefficients can provide poor coverage in some cases, even with relatively large samples



Lab Exercises
=================================================

1. What is the mean fraction of women using birth control for each age group? Each education level? For women who do or don't want more children?
     - Hint: look at the "data wrangling" cheat sheet functions `mutate`, `group_by`, and `summarize`
2. Based on ```fit1```, write on paper the model for expected probability of using birth control.  Don't forget the logit function.
3. Based on ```fit1```, what is the expected probability of an individual 25-29 years old, with high education, who wants more children, using birth control? Calculate it manually, and using `predict(fit1)`
4. Based on ```fit1```: Relative to women under 25 who want to have children, what is the predicted increase in odds that a woman 40-49 years old who does _not_ want to have children will be taking birth control?
5. Using a likelihood ratio test, is there evidence that a model with interactions improves on ```fit1``` (no interactions)?
6. Which, if any, variables have the strongest interactions?
