
---
title: "Session 8 - Survival Analysis III"
author: "Levi Waldron"
output: beamer_presentation
---

# Welcome and outline - session 8

**Learning Objectives:**

* Check model assumptions and fit of the Cox model
    + residuals analysis
    + log-minus-log plot
* Fit and interpret multivariate Cox models
    + perform tests for trend
    + predict survival for specific covariate patterns
    + predict survival for adjusted coefficients

* Explain and perform stratified analysis and its use
* Explain time-dependent covariates and when to use them
    
* Vittinghoff sections 6.2-6.4
    

# Cox proportional hazards model

* Cox proportional hazard regression assesses the relationship between a right-censored, time-to-event outcome and multiple predictors:
    + categorical variables (e.g., treatment groups)
    + continuous variables

$$
log(HR(x_i)) = log \frac{h(t|x_i)}{h_0(t)} = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ... + \beta_p x_{pi}
$$

* $HR(x_i)$ is the hazard of patient $i$ relative to baseline
* $h(t|x_i)$ is the time-dependent hazard function $h(t)$ for patient $i$
* $h_0(t)$ is the *baseline hazard function*, and is the negative of the slope of the $S_0(t)$, the baseline _survival_ function.

* Multiplicative model

# Caveats and Assumptions
* Categories with no events
    + can occur when the group is small or its risk is low
    + HRs with respect to such a reference group are infinite
    + hypothesis tests and CIs are difficult / impossible to interpret

* Assumptions of Cox PH model
    + Constant hazard ratio over time (proportional hazards)
    + Linear association between log(HR) and predictors (log-linearity) / multiplicative relationship between hazard and predictors
    + Independence of survival times between individuals in the sample

# Residuals analysis
* Residuals are used to investigate the lack of fit of a model to a given subject. 
* For Cox regression, there’s no easy analog to the usual "observed minus predicted" residual

\footnotesize
```{r}
library(pensim); set.seed(1)
mydat <- create.data(nvars=c(1, 1), nsamples=500, 
    cors=c(0, 0), associations=c(0.5, 0.5), 
    firstonly=c(TRUE, TRUE), censoring=c(0, 8.5))$data
```

```{r, echo=TRUE}
## Rename variables of simulated data, and make one variable categorical.
colnames(mydat)[1:2] <- c("Var1", "Var2")
mydat$Var1 <- cut(mydat$Var1, breaks=2, labels=c("low", "high"))
mydat$time <- ceiling(mydat$time*1000)
```

# Simulated data to test residuals methods

\footnotesize

```{r}
summary(mydat)
```

# Kaplan-Meier plot of simulated data, stratified by Var1

```{r, echo=FALSE}
##Do Kaplan-Meier plot
par(cex=1.5)
suppressPackageStartupMessages({
  library(survival)
  library(survminer)
})
kmfit <- survfit(Surv(time, cens)~Var1, data=mydat)
ggsurvplot(kmfit, risk.table = TRUE, linetype=1:2)
```

# Martingale residuals

* censoring variable $c_i$ (1 if event, 0 if censored) minus the estimated cumulative hazard function $H(t_i, X_i, \beta_i)$ (1 - survival function)
    + E.g., for a subject censored at 1 year ($c_i=0$), whose predicted cumulative hazard at 1 year was 30\%, Martingale = 0 - 0.30 = -0.30.
    + E.g. for a subject who had an event at 6 months, and whose predicted cumulative hazard at 6 months was 80%, Margingale = 1 - 0.8 = 0.2.

* Problem: not symmetrically distributed, even when model fits the data well

# Martingale residuals in simulated data
```{r, echo=FALSE}
fitsim <- coxph(Surv(time, cens) ~ Var1 + Var2, data=mydat)
par(cex=1.5)
plot(mydat$time, resid(fitsim),
     xlab="Time", ylab="Martingale Residuals")
abline(h=0, lty=2)
lines(lowess(mydat$time, resid(fitsim)), col='red')
```

# Deviance residuals in simulated data
* Deviance residuals are scaled Martingale residuals
* Should be more symmetrically distributed about zero?
* Observations with large deviance residuals are poorly predicted by the model
```{r, echo=FALSE}
par(cex=1.5)
plot(mydat$time, resid(fitsim, type="deviance"),
     xlab="Time", ylab="Deviance Residuals")
abline(h=0, lty=2)
lines(lowess(mydat$time, resid(fitsim)),col='red')
```

# Schoenfeld residuals
* technical definition: contribution of a covariate at each event time to the partial derivative of the log-likelihood
* intuitive interpretation: the observed minus the expected values of the covariates at each event time.
* a random (unsystematic) pattern across event times gives evidence the covariate effect is not changing with respect to time
* If it is systematic, it suggests that as time passes, the covariate effect is changing.

# Schoenfeld residuals for simulated data

```{r, echo=FALSE}
fitzph <- cox.zph(fitsim)
par(mfrow=c(1,2), cex=1.5)
plot(fitzph, var=1)
plot(fitzph, var=2)
```

# Schoenfeld test for proportional hazards

* Tests correlation between scaled Schoenfeld residuals and time
* Equivalent to fitting a simple linear regression model with time as the predictor and residuals as the outcome
* Parametric analog of smoothing the residuals against time using LOWESS
* If the hazard ratio is constant, correlation should be zero.
    + Positive values of the correlation suggest that the log-hazard ratio increases with time.
```{r, echo=FALSE}
fitzph
```


# The hazard function h(t), stratified by Var1

```{r, echo=FALSE}
par(cex=1.5)
library(muhaz)
hazhigh = with(mydat[mydat$Var1=="high", ], muhaz(time, cens))
hazlow = with(mydat[mydat$Var1=="low", ], muhaz(time, cens))
plot(hazlow, lwd=2, ylim=c(0, 0.002), xlab="Follow-up time")
lines(hazhigh, lty=2, lwd=2, col="red")
legend("topleft", legend=c("Var1 = low", "Var1 = high"), lty=1:2, lwd=2, col=c("black", "red"), bty='n')
```

# Log-minus-log plot

```{r, echo=FALSE}
kmfit <- survfit(Surv(time, cens) ~ Var1, data=mydat)
ggsurvplot(kmfit, fun="cloglog")
```

# Example: Primary Biliary Cirrhosis (PBC)

* Mayo Clinic trial in primary biliary cirrhosis (PBC) of the liver conducted between 1974 and 1984, n=424 patients. 
* randomized placebo controlled trial of the drug D-penicillamine.
    + 312 cases from RCT, plus additional 112 not from RCT.
* Primary outcome is (censored) time to death

# Kaplan-Meier plot of treatment and placebo arms
```{r, echo=FALSE, message=FALSE}
## Data cleaning
pbc <- read.csv("PBC.csv", row.names=1)
pbc.os <- pbc[!is.na(pbc$time) & !is.na(pbc$status) & !is.na(pbc$trt), ]
pbc.os$os <- pbc.os$status==2
pbc.os <- pbc.os[!is.na(pbc.os$status), ]
pbc.os$arm <- factor(pbc.os$trt, levels=2:1, labels=c("placebo", "treatment"))
pbc.os$ascites <- factor(pbc.os$ascites)
pbc.os$hepato <- factor(pbc.os$hepato)
pbc.os$spiders <- factor(pbc.os$spiders)
pbc.os$edema <- factor(pbc.os$edema)
pbc.os$stage <- factor(pbc.os$stage)
```

```{r, message=FALSE, echo=FALSE}
##KM plot of PBC data
par(cex=1.5)
library(survival)
fit <- survfit(Surv(time/365, os) ~ arm, data=pbc.os)
survminer::ggsurvplot(fit, risk.table = TRUE, linetype=1:2)
```


# Tests for trend

* For ordinal variables like stage (1, 2, 3, 4)
    + This is a test for linear / quadratic / cubic relationship between coefficients and their index
    + Model selection by LRT or Wald Test

```{r, echo=FALSE}
par(cex=2)
## This is a manual way to plot the trend.
fit <- coxph(Surv(time, os) ~ stage, data=pbc.os)
stagecoefs <- c(0, coef(fit)[c("stage2", "stage3", "stage4")])
plot(x=1:4, y=stagecoefs, type="b", xlab="Stages 2, 3, 4 (1=ref)", 
     ylab="Coefficient, ie ln(HR)", xlim=c(1, 4))
## In R, just define stage as an ordered factor and tests for trend are done automatically:
pbc.os$stageordered <- factor(pbc.os$stage, ordered=TRUE)
fit <- coxph(Surv(time, os) ~ stageordered, data=pbc.os)
summary(fit)  ##highly significant tests of overall fit by LRT, Wald, and logrank test.
```

# Predicted survival for specific covariate patterns
* The Cox model is a _relative_ risk model
    + only predicts relative risks between pairs of subjects 
* Key is to calculate the overall $S(t)$, then multiply it by the relative hazard for the specific covariate pattern.
* In this example we plot the baseline survival for all stages together, then for stages 1-4 separately. 

# Predicted survival for specific covariate patterns

```{r, echo=FALSE}
par(cex=1.5)
fit <- coxph(Surv(time, os) ~ stage, data=pbc.os)
##HR for each stage, relative to overall
HR <- predict(fit, newdata=data.frame(stage=factor(1:4)))
plot(survfit(fit)$time, survfit(fit)$surv, type="l", xlab="Time (days)", ylab="Survival Probability")
for (i in 1:4)
  lines(survfit(fit)$time, (survfit(fit)$surv)^exp(HR[i]), type="l", col=i+1, lty=i+1)
legend("bottomleft", legend=c("Overall", "Stage 1", "Stage 2", "Stage 3", "Stage 4"), col=1:5, lty=1:5, bty="n")
```

# Multivariate regression

* Same coding and objectives as for `lm()` and `glm()`
    + controlling for confounding
    + testing for mediation
    + testing for interaction

# Multivariate regression

\footnotesize
```{r}
fit <- coxph(Surv(time, os) ~ age + sex + edema 
                + stage + arm, data=pbc.os)
summary(fit)
```

# Predicted survival for adjusted coefficients

* Can create Kaplan-Meier curves for crude or unadjusted coefficients
    + Section 6.3.2.3 in Vittinghoff
* Idea is to estimate hazard ratio in an unadjusted model:

\footnotesize
```{r}
unadjfit <- coxph(Surv(time, os) ~ stage, data=pbc.os)
coef(unadjfit)
```

# Predicted survival for adjusted coefficients

* and in an adjusted model:

\footnotesize
```{r}
adjfit <- coxph(Surv(time, os) ~ age + sex + edema 
                + stage + arm, data=pbc.os)
coef(adjfit)
```

# Predicted survival for adjusted coefficients (cont'd)

* The survival function will be calculated for a "baseline" group, say stage 1, then exponentiated with the adjusted coefficient, e.g.:
$$
[S_{stage=1}(t)]^{exp(\beta_{stage=4})}
$$

```{r, echo=FALSE}
par(cex=1.5)
basefit <- survfit(unadjfit, newdata=data.frame(stage=factor(1)))
plot(x=basefit$time, y=basefit$surv, type="l", ylim=c(0, 1), xlab="Time", ylab="Survival Probability")
stage4unadj <- (basefit$surv)^(exp(coef(unadjfit)["stage4"]))
stage4adj <- (basefit$surv)^(exp(coef(adjfit)["stage4"]))
lines(x=basefit$time, y=stage4adj, col="red", lty=1, lw=2)
lines(x=basefit$time, y=stage4unadj, col="blue", lty=3)
legend("bottomleft", legend=c("Stage 1", "Stage 4 adjusted", "Stage 4 unadjusted"),
       col=c("black", "red", "blue"), lw=c(1,2,1), lty=c(1,1,3))
```

# Stratification

* Vittinghoff 6.3.2
* Separates the analysis into strata
    + must have an adequate number of events in each stratum (at least 5 to 7)
    + can be used to adjust for variables with strong impact on survival
    + can help solve proportional hazards violations
    
* Strata have different baseline hazards
* Coefficients / Hazard Ratios are calculated within stratum then combined.

# Stratification

**Example - in R, strata() can be added to any model formula**
\footnotesize
```{r}
mycox <- coxph(Surv(time, os) ~ trt + strata(stage), data=pbc.os)
summary(mycox)
```

# Immortal Time Bias in observational studies

* For example, Yee *et al.* reported that new statin users reported a 26% reduction in the risk of diabetes progression with one year or more of treatment relative to never-users (adjusted HR 0.74, 95% CI: 0.56 to 0.97).
    + New users excludes those who had received a lipid lowering drug from three years before to six months after cohort entry
* This is a surprising finding because of confounding: people whose diabetes progresses are more likely to develop cardiovascular disease, an indication for statins.
    + would result in HR > 1
* Why?

\tiny

* Yee *et al.* Statin use in type 2 diabetes mellitus is associated with a delay in starting insulin (http://onlinelibrary.wiley.com/doi/10.1111/j.1464-5491.2004.01263.x/full)

# Immortal Time Bias in observational studies

* Why?
    + all person days between cohort entry and end of follow-up were **classified as treated** for those who met the statin user definition, regardless of the date on which they met this definition and as untreated for non-users
    + thus all persons in the *treated* group are "immortal" from time 0 until the initiation of statin treatment
    + this period of immortality makes treatment look more effective

# A solution: Time-dependent covariates (TDC)

**Definition: A time-dependent covariate in a Cox model is a predictor whose values may vary with time.**

* A solution is treating statin prescription as a Time Dependent Covariate (TDC) (Levesque 
*et al.*)
* Levesque *et al.* Problem of immortal time bias in cohort studies: example using statins for preventing progression of diabetes (https://doi.org/10.1136/bmj.b5087)


# Lab exercises

Use the **PBC** dataset for these exercises.

1. Does the dataset actually contain `n=424` patients as stated above?
2. For how many patients is there complete data for `time`, `status`, and `trt`?
3. Which variables are categorical, and which are continuous?
4. Make a Kaplan-Meier plot for overall survival, stratified by the `trt` variable. 
5. Fit univariate Cox models for each available covariate, using a loop. Which have significant p-values (you can ignore multiple testing for now)?
6. Fit a multivariate Cox model with `trt` and `spiders` as covariates. 
     a. Interpret the coefficients and p-values from this multivariate model. 
     b. Create a log-minus-log plots for `treatment+spiders` model. At what times, if any, does it look like there might be any violation of the proportional hazards assumption?
7. Fit a Cox model with `trt` as a covariate while stratifying by `spiders`. Does this stratification make any relevant difference?
8. Suppose it is hypothesized that the treatment effect is time-dependent, and changes at 5 years. Create a time-dependent covariate that is a step function at 5 years, and test this hypothesis by using an interaction term between the new time-dependent covariate and the `trt` variable, with these two covariates plus `spiders`, in the model.
