---
title: "MDS exercises"
author: "Levi Waldron"
date: "March 10, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tissuesGeneExpression)
data(tissuesGeneExpression)
head(e)
head(tissue)
```

From the previous "Distance and Dimension Reduction" exercise:
```{r, cache=TRUE}
s = svd(e - rowMeans(e))
z = s$d*t(s$v)
```

# What is the absolute value of the correlation between the first dimension of z and the first dimension in mds?

```{r}
d = dist(t(e))
mds = cmdscale(d)
cor(mds[, 1], z[, 1])
```

# What is the absolute value of the correlation between the second dimension of z and the second dimension in mds?

```{r}
cor(mds[, 2], z[, 2])
```

# Which dimension of z most correlates with the outcome sampleInfo$group ?

```{r}
library(GSE5859Subset) 
data(GSE5859Subset)
```

```{r}
s = svd(geneExpression-rowMeans(geneExpression)) 
z = s$d * t(s$v)
```

```{r}
res <- apply(z, 2, function(pc) cor(pc, sampleInfo$group))
names(res) <- 1:ncol(z)
sort(res)
```

# What is this max correlation?

```{r}
max(res)
```

# Which dimension of z has the second highest correlation with the outcome sampleInfo$group?


